Viral and Bacterial Communities of Colorectal Cancer

### 2017-03-26 doc/manuscript.Rmd

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the majority of colorectal cancer case causes remain unclear, its progression has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community, but have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities. Together our findings add to the existing model for the microbiome's role in colorectal cancer development, thus providing a more complete understanding of colorectal cancer etiology.

**Word Count**: 226 / 250

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its progression and severity have been linked to colonic bacterial community composition. Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compared to bacterial community changes. The results suggested an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer, as well as provide a new understanding of basic colorectal cancer etiology.

**Word Count**: 113 / 120

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer), and has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer together suggest that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo]. Although the majority of colorectal cancer case causes remain unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

@@

Here we begin addressing the knowledge gap in whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity. We also aim to evaluate their diagnostic potential. We report that colorectal cancer is associated with variation in colonic virus community composition. The majority of viruses identified within the virome were temperate bacteriophages. Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was driven by a subset of influential phages. Our data suggest that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics. To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.

We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after subsampling, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Overall, standard diversity metrics were insufficient for capturing virus community differences between disease states.

## Altered Virome Composition in Colorectal Cancer
16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals. These models were based on virus metagenomic community relative abundance profiles and were compared to 16S rRNA gene models. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. Statistically, the viral model performed better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), although the results were practically similar, with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the synergistic capabilities of the bacteria and viruses, we built a combinatory model that used both bacterial and viral community data. The combination model yielded statistically significant but minor improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed more poorly than the viral model, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This performance was also much poorer than that observed in the 16S rRNA gene model. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**. Filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.

The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven by a few operational genomic units, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome. This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Shifted Phage Influence Across Cancer Stages
Because our cohort included healthy, adenomatous, and cancerous colons, we were able to evaluate shifts in influential phage identities during cancer progression. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions. While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).

The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.

The classification model determined cancer state by incorporating the relative abundance profiles of the microbes within each community. Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**. Not all important OGUs were of increased abundance in the diseased state. The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.

@@

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings are consistent with previous reports suggesting the gut virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

@@

Given these findings, we hypothesized that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network. We investigated the infectious capabilities and potential host ranges of all phage OGUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**). The predicted interactions were then used to identify community hub phages. This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. The phages with high centrality values were defined as community hubs. We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

@@

Our data allowed us to begin delineating the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**. We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusive predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundance of these entities. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.

>>>>

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although etiology for the majority of colorectal cancer cases remain unclear, its progression has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community. However, they have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

**Word Count**: 201 / 250

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its progression and severity have been linked to colonic bacterial community composition. Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We addressed this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients and how they compared to bacterial community changes. The results suggested an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer, as well as provide a new understanding of basic colorectal cancer etiology.

**Word Count**: 111 / 120

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer).[Comment: the gut virome isn't associted with the following diseases...] The human virome has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggests that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans died from the disease [@Siegel:2014jo]. Although the etiology for [cite specific number - 80%?] of colorectal cancer cases remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

@@

Here we address the gap in knowledge of whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity. We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers. We report that colorectal cancer is associated with variation in colonic virus community composition. The majority of viruses identified within the virome were temperate bacteriophages. Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was driven by a subset of influential phages. Our data suggest that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer. [Comment: I worry that from "We report..." to the end of the paragraph is too much a rehash of the Results. I would suggest moving this section to the abstract since the abstract is a bit light on results. ]

@@

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained DNA at a minimal concentration (XX ng/ul), thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
[Comment: What do you think about using operational viral units (OVU) rather than OGU? I think of OGU as representing bacterial genomes. Adding phage to the name would make it more clear what we are refering to]

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics. ~~To control for uneven sequencing depths, we rarefied to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.~~ [Comment: put this in the methods section rarefied != sub-sampled, make sure you rarefied (i.e. subsampled 1000 times)] We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Similar to what was previously seen when analyzing the 16S rRNA gene sequences from the same samples [@Zackular:2014fba], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.

## Altered Virome Composition in Colorectal Cancer
16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. Statistically, the viral model performed better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), although the results were practically similar, with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

[Comment:
1A: Get rid of the legend title "Disease". I feel like the 0.1 and 0.9 marks on the ROC curves adds clutter without improving interpretation.
1B: Can you order: Bacteria, Virus, Combined, Metagenome - why is the combined AUC > virus when it seems the opposite in A? Can you put the AUCs in the legend for 1A and get rid of 1B?
1C: Format OTU names to be like "OTU 4" rather than "Otu00004"
1D: Format cluster  names to be like "Cluster 115" rather than "Cluster_115"
1E: See comments for 1C and 1D
Can you justify the y-axis in C, D, and E so they are vertically aligned and the x-axes are the same lengths?
]

To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model yielded statistically significant but small performance improvement beyond the viral (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that features from the virus and bacterial communities interact to improve the ability to classify stool as belonging to individuals with cancerous colons.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed more poorly than any of the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**. [Comment: Can you add a comment about the nature of the features in the viral model?]

The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven by a few OGU, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome. This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Shifted Phage Influence Across Cancer Stages
Because our cohort included people with healthy, adenomatous, and cancerous colons we were able to evaluate shifts in influential phage identities during cancer progression. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. {Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**. Not all important OGUs were of increased abundance in the diseased state. The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.} [Comment: I don't feel like the text between the {}'s says anything. It's like word salad.]

[Comment: Can't distinguish colors in Fig S9F (and axis label overlaps panel label)]

@@

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses indicating that the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

@@

[ Comment: how are the rows and columns sorted in Figure 2? Looks like by importance, how about sorting them by co-occurrence? ]

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network. We investigated the infectious capabilities and potential host ranges of all phage OGUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**). The predicted interactions were then used to identify community hub phages. This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. The phages with high centrality values were defined as community hubs. Phage OGU centrality was significantly and positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

@@

Our data allowed us to begin delineating the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**. We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusive predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter(the Microbiome article)]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.

In addition to modeling the basic potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu]. We hypothesize that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria (such as *Fusobacterium nucleatum*) were able to colonize. Once the initial influential bacteria had established themselves in the epithelium, other opportunistic bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm. Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria were then able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundance was most influential. As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could have continued mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria. In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role of the virome in colorectal cancer.
